He-Man: A domain specific language for lightweight concurrency

### 2012-03-11 paper/paper.tex
A lot of editing.

\documentclass[preprint]{sigplanconf}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
{hyperref}
\usepackage{graphicx}
\usepackage{verbatim}

\usepackage{listings}
\lstset{
  language=Haskell,
  basicstyle=\small\ttfamily,
  frame=trbl,
  basewidth={0.5em,0.45em},
%  commentstyle=\ttfamily\color{ForestGreen},
  literate=*{+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
            {>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
            {\\\\}{{\char`\\\char`\\}}1
            {->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
            {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
            %{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
            {>>}{{>>}}2 {>>=}{{>>=}}2
            {|}{{$\mid$}}1
}

\lstnewenvironment{egcode}
  {\lstset{frame=TRBL}}
  {\lstset{frame=trbl}}

\lstnewenvironment{code}
  {\lstset{frame=}}
  {}

\newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
\renewcommand{\t}{\texttt}
\renewcommand{\b}{\textbf}
\renewcommand{\i}{\textit}

\newcommand{\prettybox}[1]{\fbox{\parbox{\linewidth}{#1}}}

% A bunch of crazy BNF TeX I wrote once
\def\bnf#1{\prettybox{\halign{##\hfil& \hfil ##\hfil& \ ##\hfil\cr #1}}}
\def\p#1{& = & #1 \cr}
\def\pp#1{& & #1 \cr}
\def\a#1{& $\mid$ & #1 \cr}

% XXX: We need to cite ElliottEtAl somewhere!
% XXX: standardize on IO vs I/O
% XXX: standardize on domain-specific vs domain specific

@@

% TODO: talk about type system, runtime improvements, optimizations

@@

Network applications require a very constrained form of
concurrency. In most network applications, I/O (both network
and disk) latency dominates computation time, so each client's thread
of execution spends most of its time waiting on I/O. In particular, we
would like to always yield to a client no longer blocked on I/O; we
need not yield to blocked clients, nor do we (usually) need arbitrary
interleaving of execution.

In the event loop model, network I/O is done using non-blocking
\t{read} and \t{write} calls and all of the blocking for I/O is
centralized in a single call to a multiplexing system call such as
\t{select} or \t{epoll}. The multiplexing call will block until some 
I/O is ready to be performed, and indicates to the application
which file descriptors are ready. To decide how to handle these 
events, an application must, for each connection, have some sort of 
state machine to track what state the connection is in and what 
needs to happen next, as well as other local state.

One common scheme for doing this involves registering a function
pointer with the event loop to be called on an event. Typically each
connection is also associated with a structure of connection local
data. Instead of making a blocking system call, then, an application
registers an event with the event loop along with the function to call
when it occurs, and then returns back to the loop.

The advantage of this scheme is that it is highly scalable and can be
efficient: handling a new connection typically just involves
allocating a new structure and registering it with the event loop. The
disadvantage is that now network facing code needs to be scattered
through many small functions, with all access to connection state
mediated through a shared structure.

The event loop model is then, essentially, built around closure-
converted trampolining continuation passing style. These are well
studied problems and can easily be done mechanically in the He-Man
backend.

@@

It is important to provide some sort of type discipline for He-Man
programs. Providing a type system in the front end allows us to
guarantee that the C we output does not contain type errors, vastly
simplifying development. We also take advantage of this to provide
type abstractions above and beyond C's limited type system. % XXX: bad wording
For example, He-Man's type system contains an abstract
$\b{FD}$ type that is distinct from the type of integers.

We offload the work of type checking He-Man programs to the Haskell type
checker. Our method for doing this is fairly standard, following Leijen and
Meijer \cite{LeijenMeijer} in using phantom types to enforce typing.

A fragment of the code implementing the He-Man type system is shown in
Figure \ref{fig:types}. The internal type and expression
representations are the datatypes \t{IExpr} and \t{IType}, and
typed expressions are represented with the \t{Expr} data type
constructor, which is a trivial wrapper around \t{IExpr} parameterized
over a phantom type. Operations on typed expressions have type
signatures that restrict them to the proper types, and then peel away
the wrapper to work on the \t{IExpr} underneath.

% XXX: say something about GADTs??
Types are represented by the \t{Type} GADT that is also parameterized
over a phantom type representing the underlying type. Expression
builder functions will convert \t{Type} values are to \t{Itype} using
the \t{toItype} function, which does a simple pass-through conversion.
The decision to make \t{Type} a GADT instead of a simple wrapper with
constructor functions in the same manner as \t{Expr} is for purely
aesthetic reasons: we felt that type annotations in He-Man code were
more clearly readable if they were capitalized, which required that
they be datatype constructors.
\footnote{The authors believe that this may be the worst reason to
  choose to use GADTs to date.}

@@

Type checking foreign function calls and thread declaration and
spawning is somewhat more interesting.  For foreign function calls,
our main concern was being able to write a single wrapper for call
that handles a varying number of typed arguments. We want more than
this, however: we want to ensure that both the code for a declared
thread uses its arguments properly at the declared types code and that
spawning threads provides arguments of the correct types.
% and maybe a bit novel? I dunno.

A fragment of the code implementing typing for these features is shown
in Figure \ref{fig:arg-packets}. The key is the multi-parameter type
class \t{ArgPacket}. \t{ArgPacket} is parameterized over related
tuples \footnote{Technically speaking, the unary \t{ArgPacket}s are
  not actually tuples.} of expressions and types. The idea is that
there exists an instance \t{ArgPacket exps tys} if and only if
\t{exps} and \t{tys} are tuples of \t{Expr}s and \t{Type}s,
respectively, the tuples have the same number of elements, and for
each pair of corresponding elements in the two tuples, the \t{Type}
represents the type of the \t{Expr} (that is, they have the same
phantom type argument). To enforce this one-to-one correspondence and
to allow the Haskell inferencer to use the information it provides, we
use functional dependencies % XXX: cite, or something?
to declare that both \t{exps} and \t{tys} are uniquely determined by
the other.

Then, by parameterizing the wrapper functions for foreign function
calls and thread declaration and spawning over \t{ArgPacket exps tys},
we get the desired typing guarantees.
% XXX: is this explained fully?

@@

The He-Man runtime is written in C and provides a state machine based
thread scheduling model. In the view of the runtime, a thread consists
of some bookkeeping state, some private state, and the next
continuation function the thread is to run. The main part of the
runtime's state consists of the runqueue, which contains a list of all
currently runnable threads.

Continuation functions take a pointer to the thread information,
perform some computation (such as making non-blocking I/O calls),
update the continuation field in the thread data, and return whether
the thread is still runnable; to block, a continuation function
registers itself as blocking on an event and returns false; to exit,
it frees the thread structure and returns false.  Running a thread,
then, consists of repeatedly invoking the thread's continuation
function until it returns false. To prevent a thread that does not
need to block from starving the rest of the system, after some number
of invocations of the continuation (currently 50), the thread is
paused and put on the back of the runqueue.

The He-Man runtime main loop is very simple, then: each worker thread
loops indefinitely, and on each iteration it removes the thread at the
head of the runqueue (if it exists) and runs it, and polls for I/O
events. A global runqueue is shared between threads; a more clever
runtime could have per-worker-thread runqueues and do some sort of
work stealing.

The runtime provides a blocking mechanism based on an event
abstraction. To the runtime, an event consists of a type tag, a
pointer to a thread that is blocked on it and some event type specific
data (such as a file descriptor or a pointer to an associated
channel). Events are level-triggered: blocking on an event will return
whenever the event ``is ready'', as opposed to when the event
``becomes ready'' as in edge-triggered (using edge-triggered semantics
causes extra difficulties in an environment with multiple threads
processing events).

@@

The actual polling for I/O events is done by calling \t{epoll}
(although it would not be difficult to adapt it to use BSD's
\t{kqueue} or the more standard \t{poll} or \t{select}).  For each
event that has been returned by \t{epoll}, if a thread is waiting on
the event, the thread is added to the runqueue. If the the runqueue is
empty, the call to \t{epoll} blocks, otherwise it is called with a
timeout of zero seconds, to simply poll for events.

@@

To evaluate He-Man, we built several network applications with He-Man
and compared them to similar network applications built in C.  We also
discuss He-Man's memory overhead.

@@

To compare the ease of writing event based network applications, we
built simple echo servers using both He-Man and C. The C version was
written in as an event loop driven explicit state machine using the
He-Man runtime. The C version, which is fairly tightly written and
which uses the He-Man runtime in its capacity as a somewhat generic
event based IO framework, is 98 lines of source, according to
SLOCCount. It is fairly readable, but its structure is obscured by
being split into states and the need to explicitly handle read and
write returning \t{EAGAIN} and write returning short. The state
machine has a total of five states. The He-Man version is 33 lines
long and is a very clear representation of the control flow of each
individual thread. (Additionally, the He-Man library has 25 lines of
code implementing generic routines for simulating blocking IO using
nonblocking IO and wait.) The total line counts, however, somewhat
understate the difference between the versions, since the boilerplate
for setting up connections is about the same for both versions. When
just looking at the core read/write loop, the difference is much more
apparent: the C version is 38 lines of code, and the He-Man version is
5.

The generated code is substantially less tight than the C version,
although it is unlikely to be any slower in practice. It has 10 states
and is 240 lines of C. The echo server was able to handle 10,000
concurrent, occasionally transmitting, connections with no
problem. With 10,000 clients connected the echo server used about 40
MB of memory, which is almost entirely accounted for 4 KB buffer that
each connection maintained for I/O.

@@

To test the suitability of our system for more complicated programs
and to demonstrate separating network logic and application logic, we
built a very simple HTTP server. The network logic is written in
He-Man and request parsing and response header building is written in
C. The C request parsing is implemented as a function that takes a
buffer and returns whether it forms a complete HTTP request and, if
so, fills a buffer with the file name to serve. The entire server is
68 lines of He-Man Haskell and 39 lines of C. It suffers from a number
of major deficiencies, but is complete enough to benchmark. To make
sure that it does not get away with sending less data than the servers
it is being compared against, it sends some ``padding'' headers with
useless data.

Our benchmark measurements are taken using weighttp, an http
benchmarking tool associated with lighttpd
\cite{weighttp}. Measurements were collected by requesting a file many
times with a variable number of concurrent connections; the experiment
was repeated with a 100 byte file (80000 requests), a 16 kilobyte file
(30000 requests), and a 128 kilobyte file (8000 requests).
weighttp was run with 4 threads making requests.
Since the test repeatedly requests the same file, it should always be in
the kernel's disk cache.

For our experiments, both the client and server ran Linux kernel
2.6.32 on quad core 3.1 GHz Intel Core i5-2400 processors with 8
gigabytes of RAM. The machines were connected over a 100 megabit
Ethernet network.
We ran the tests against our He-Man HTTP server, Apache 2.2.14 (installed
with the default Ubuntu 10.04 LTS settings, except for client, thread,
and process limits being adjusted upwards) and Lighttpd 1.4.26
(installed with the default Ubuntu 10.04 LTS settings).

Each trial was run 10 times; the plots show the arithmetic mean of the
number of requests per second the servers handled; error bars show the
standard deviation.

@@

The per-thread memory overhead imposed by He-Man is very small: it
consists of a thread id, list pointers for the scheduling queue, the
next state to enter, a pointer to the most recently received event,
and list heads and tails for lists of events owned by the thread and
buffers owned by the thread (the latter two are so that we can
automatically reclaim these resources when the thread exits). On a
64-bit system, the total overhead comes out to 64 bytes, and could
easily be reduced further (e.g., by using singly-linked lists).  For
comparison, the default thread stack size under NPTL on Linux is 2
megabytes.

% see how many threads we can spawn with Memory.hs

% Haskell EDSLs:
%\cite{Gill}
%\cite{ElliottEtAl}
%\cite{LeijenMeijer}

@@

@@

XXX: I'm not really sure what we want to put here. I think maybe we
shouldn't have a future work section. If we just want to talk about
the deficencies of the thing we built, that should go somewhere else,
maybe? It isn't really clear we have obvious future research
directions to talk about.

>>>>

\documentclass[preprint]{sigplanconf}
\usepackage[colorlinks=true,linkcolor=black,urlcolor=black,citecolor=black]%
{hyperref}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}

\lstset{
  language=Haskell,
  basicstyle=\small\ttfamily,
  frame=trbl,
  basewidth={0.5em,0.45em},
  commentstyle=\ttfamily\color{ForestGreen},
  literate=*{+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
            {>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
            {\\\\}{{\char`\\\char`\\}}1
            {->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
            {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
            %{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
            {>>}{{>>}}2 {>>=}{{>>=}}2
            {|}{{$\mid$}}1
}
\lstnewenvironment{code}{\lstset{frame=}}{}

\newcommand{\prettybox}[1]{\fbox{\parbox{\linewidth}{#1}}}
\newenvironment{smallverbatim}{\endgraf\small\verbatim}{\endverbatim} 
\renewcommand{\t}{\texttt}
\renewcommand{\b}{\textbf}
\renewcommand{\i}{\textit}

% For BNF grammars:
\def\bnf#1{\prettybox{\halign{##\hfil& \hfil ##\hfil& \ ##\hfil\cr #1}}}
\def\p#1{& = & #1 \cr}
\def\pp#1{& & #1 \cr}
\def\a#1{& $\mid$ & #1 \cr}

% XXX: standardize on IO vs I/O
% XXX: standardize on continuation-passing vs continuation passing
% XXX: standardize on domain-specific vs domain specific
% TODO: talk about type system, runtime improvements, optimizations

@@

@@

Network applications require a very constrained form of
concurrency. In most network applications, network and disk
I/O latency dominate computation time, so each client's thread
of execution spends most of its time waiting on I/O. In particular, we
would like to always yield to a client no longer blocked on I/O; we
need not yield to blocked clients, nor do we (usually) need arbitrary
interleaving of execution.

In the event-driven model, network I/O is performed with non-blocking
\t{read} and \t{write} calls, and all blocking on I/O is consolidated
in a single, centralized call to a multiplexing system call such as
\t{select} or \t{epoll}. This multiplexing call blocks until some 
I/O is ready to be performed, and indicates to the application
which file descriptors are ready. To decide how to handle these 
events, an application must, for each connection, maintain a
state machine to track the connection's current and next states, as 
well as any local data.

One common solution involves registering a function
pointer with the event loop to be called on an event. Typically, each
connection is also associated with a structure of connection-local
data. Instead of making a blocking system call, an application
registers an event with a function pointer to call once the event 
occurs, and then returns back to the loop.

The advantage of this scheme is that it is highly scalable and can be
efficient: handling a new connection requires only allocating a new 
structure and registering it with the event loop. The disadvantage is 
that network-facing code must be scattered across many small functions 
which call each other only through function pointers, and all connection 
state is accessed through a shared structure.

However, as we discuss in the next section, these code transformations are
familiar to functional programmers as closure conversion (to access connection
state across function boundaries), continuation-passing style (via function
pointers) with trampolining (yielding to the event loop after registering I/O).
These are well-studied conversions which can be performed mechanically by the
He-Man backend.

@@

We provide a simple type system for He-Man programs which guarantees that we
generate C code which is free of type errors. We take advantage of this to
provide abstractions not present in C, e.g., the \b{FD} type of file descriptors
is distinct from integers in He-Man, while C draws no distinction between them.

We type-check He-Man programs with the Haskell type checker, following Leijen
and Meijer \cite{LeijenMeijer} in using phantom types. A fragment of the code
implementing the He-Man type system is shown in Figure \ref{fig:types}. Types
and expressions are internally represented by the datatypes \t{IExpr} and
\t{IType}, and typed expressions are represented with the \t{Expr} data type
constructor, which is a trivial wrapper around \t{IExpr} parameterized
over a phantom type. Operations on typed expressions have type signatures that
restrict them to the proper types, and discard the wrapper to act on the
\t{IExpr} underneath.

Types are represented by the \t{Type} GADT, which is parameterized over a
phantom type representing the underlying type. When building expressions, we
convert \t{Type} values to \t{IType} with the \t{toItype} function. The decision
to make \t{Type} a GADT instead of a simple wrapper with constructor functions
in the same manner as \t{Expr} is purely aesthetic: we felt that type
annotations in He-Man code were more clearly readable if they were capitalized,
which required them to be datatype constructors.
%\footnote{The authors believe that this may be the worst reason to
%  choose to use GADTs to date.}

@@

Type-checking foreign function calls and thread declaration and spawning is
somewhat more interesting. We wanted a single wrapper for foreign function calls
to handle a varying number of typed arguments; we also wanted to ensure that
thread spawn sites pass in values of the correct types, and declared threads use
their arguments in type-safe ways.

A fragment of the code implementing typing for these features is shown
in Figure \ref{fig:arg-packets}. The key is the multi-parameter type
class \t{ArgPacket}. \t{ArgPacket} is parameterized over related
tuples\footnote{Technically speaking, the unary \t{ArgPacket}s are
not actually tuples.} of expressions and types. The idea is that
there exists an instance \t{ArgPacket exps tys} if and only if
\t{exps} and \t{tys} are tuples of \t{Expr}s and \t{Type}s,
respectively; the tuples have the same number of elements; and for
each pair of corresponding elements in the two tuples, the \t{Type}
represents the type of the \t{Expr} (that is, they have the same
phantom type argument).

We use functional dependencies to declare that both \t{exps} and \t{tys} are
uniquely determined by the other, which enforces this one-to-one correspondence
and allows the Haskell inferencer to use provided typing information. We obtain
our desired typing guarantees by parameterizing the wrapper functions for
foreign function calls and thread declaration/spawning over \t{ArgPacket exps tys}.

@@

The He-Man runtime is a multithreaded C program which provides a generic state
machine thread scheduling model. To the runtime, a thread consists of some
bookkeeping state, some private state, and the thread's next continuation
function. The runtime's state is primarily in the runqueue, which contains a
list of all currently runnable threads.

Each continuation function takes a pointer to the thread information,
performs some computation (e.g., making non-blocking I/O calls),
updates the continuation field in the thread data, and returns whether
the thread is still runnable. To block, a continuation function
registers itself as blocking on an event and returns false; to exit,
it frees the thread structure and returns false. Running a thread,
then, consists of repeatedly invoking the thread's continuation
function until it returns false. To prevent a thread that does not
need to block from starving the rest of the system, after a continuation is
invoked some number of times (currently 50), the thread is paused and moved to
the back of the runqueue.

The He-Man runtime main loop is very simple: each worker thread
loops indefinitely, and on each iteration it removes the thread at the
head of the runqueue (if it exists) and runs it, and polls for I/O
events. A global runqueue is shared between threads; a fancier 
implementation could maintain per-worker-thread runqueues and use
work stealing.

The runtime provides a blocking mechanism based on an event
abstraction. To the runtime, an event consists of a type tag, a
pointer to a thread that is blocked on it and some event type specific
data (such as a file descriptor or a pointer to an associated
channel). Events are level-triggered: blocking on an event will return
whenever the event ``is ready'', not when the event ``becomes ready''
(edge-triggered). We make this design decision because edge-triggered semantics
cause difficulties in an environment where multiple threads are processing
events.

@@

The actual polling for I/O events is done by calling \t{epoll}, and could be
easily adapted to BSD's \t{kqueue}, or the more standard \t{poll} or \t{select}.
For each event that \t{epoll} returns, if a thread is waiting on that event, the
thread is added to the runqueue. If the the runqueue is empty, the call to
\t{epoll} blocks; otherwise it is called with a timeout of zero seconds, to
simply poll for events.

@@

To evaluate He-Man, we built several network applications with He-Man
and compared them to similar network applications built in C.

@@

To compare the ease of writing event-driven network applications, we wrote
simple echo servers in both He-Man and C. The C version was written as an
explicit state machine using the He-Man runtime in its capacity as a generic
event-driven I/O framework.  This code is fairly tightly written, and consists
of 98 lines of source code, according to SLOCCount.  It is fairly readable, but
its structure is obscured by being split into states, and the need to explicitly
handle read and write returning \t{EAGAIN}, or write returning short. The state
machine has a total of five states.

The He-Man version is 33 lines long and clearly represents the control flow of
each individual thread. (This does not include the He-Man library, which has 25
lines of code implementing generic routines for simulating blocking IO using
nonblocking IO and wait.) The total line counts, somewhat understate the
difference between the versions, since the boilerplate for setting up
connections is about the same for both versions. When just looking at the core
read/write loop, the difference is much more apparent: the C version is 38 lines
of code, and the He-Man version is 5.

The generated code is substantially longer than the C version, although not any
slower in practice; it has 10 states in 240 lines of C. This echo server easily
handles 10,000 concurrent, occasionally transmitting, connections.  With 10,000
clients, the echo server uses about 40 MB of memory, almost entirely consisting
of a 4 KB I/O buffer for each connection.

@@

To test the suitability of our system for more complicated programs and to
demonstrate separating network logic from application logic, we also built a
very simple HTTP server. The network logic is written in He-Man; request parsing
and response header building are implemented in C. Request parsing is handled by
function which takes a buffer and returns whether it forms a complete HTTP
request; if so, it fills a buffer with the file name to serve. The entire server
is 68 lines of He-Man and 39 lines of C. Although it is very incomplete as an
HTTP server, it is sufficient for benchmarking purposes. We add some useless
headers for padding, to ensure that it sends the same amount of data as the
complete servers we benchmark against.

Our benchmark measurements are taken using weighttp \cite{weighttp}, an http
benchmarking tool associated with lighttpd.  Measurements were collected by
requesting a file many times with a variable number of concurrent connections;
the experiment was repeated with a 100 byte file (80,000 requests), a 16 KB file
(30,000 requests), and a 128 KB file (8,000 requests). weighttp was run with 4
threads making requests. Since the test repeatedly requests the same file, it
should always be in the kernel's disk cache.

For our experiments, both the client and server ran Linux kernel 2.6.32 on quad
core 3.1 GHz Intel Core i5-2400 processors with 8 gigabytes of RAM. The machines
were connected over a 100 megabit Ethernet network. We ran the tests on our
He-Man HTTP server, Apache 2.2.14 (installed with the default Ubuntu 10.04 LTS
settings, except for client, thread, and process limits being adjusted upwards),
and Lighttpd 1.4.26 (installed with the default Ubuntu 10.04 LTS settings).

Each trial was run 10 times. The plots show the arithmetic mean of the number of
requests per second handled by each server; error bars show the standard
deviation.

% XXX work on the below paragraphs

@@

The per-thread memory overhead imposed by He-Man is very small: it consists of a
thread id, list pointers for the scheduling queue, the next state to enter, a
pointer to the most recently received event, and list heads and tails for lists
of events owned by the thread and buffers owned by the thread.  (The event lists
allow us to automatically reclaim resources when a thread exits). On a 64-bit
system, the total overhead comes out to 64 bytes, and could be reduced further.
For comparison, the default thread stack size under NPTL on Linux is 2 MB.

@@

% Haskell EDSLs:
%\cite{Gill}
%\cite{ElliottEtAl} CITE THIS
%\cite{LeijenMeijer}

@@
